Gut-resident microorganisms and their genes are associated with cognition and neuroanatomy in children

### 2023-10-19 manuscript/main.tex
update figure and table legend titles

\begin{table}[!h]
    \centering
    \begin{tabular}{|r|r|r|r|r|r|}
      \hline\hline
      \textbf{Group}              & \textbf{Subgroup}        & \textbf{All} & \textbf{Under 6 months} & \textbf{Over 18 months} & \textbf{Future} \\\hline
      N subjects                  &                          & 381 & 82 & 277 & 104 \\  \hline
    \multirow{3}{*}{Samples}      & 1                        & 214 (56.2\%) &  &  &  \\ \cline{2-6}
                                  & 2                        & 90 (23.6\%) &  &  &  \\ \cline{2-6}
                                  & \textgreater 2           & 77 (20.2\%) &  &  &  \\ \hline
    \multirow{3}{*}{Age (months)} & min                      & 1.3 & 2.89 & 18.03 & 1.3 \\ \cline{2-6}
                                  & max                      & 119.53 & 5.95 & 119.53 & 11.95 \\ \cline{2-6}
                                  & median                   & 24.62 & 3.64 & 41.59 & 6.08 \\ \hline
    \multirow{2}{*}{Sex}          & F                        & 172 (45.1\%) & 42 (51.2\%) & 122 (44.0\%) & 47 (45.2\%) \\   \cline{2-6}
                                  & M                        & 209 (54.9\%) & 40 (48.8\%) & 155 (56.0\%) & 57 (54.8\%) \\  \hline
    \multirow{6}{*}{Race}         & White                    & 265 (69.6\%) & 48 (58.5\%) & 201 (72.6\%) & 62 (59.6\%) \\   \cline{2-6}
                                  & Black                    & 40 (10.5\%) & 15 (18.3\%) & 21 (7.6\%) & 17 (16.3\%) \\ \cline{2-6}
                                  & Indigenous               & 1 (0.3\%) & 0 (0.0\%) & 1 (0.4\%) & 0 (0.0\%) \\ \cline{2-6}
                                  & Asian                    & 8 (2.1\%) & 2 (2.4\%) & 7 (2.5\%) & 3 (2.9\%) \\ \cline{2-6}
                                  & Mixed                    & 54 (14.2\%) & 14 (17.1\%) & 38 (13.7\%) & 16 (15.4\%) \\ \cline{2-6}
                                  & Other                    & 3 (0.8\%) & 1 (1.2\%) & 3 (1.1\%) & 2 (1.9\%) \\ \hline
    \multirow{6}{*}{Maternal Ed.} & Junior high school       & 2 (0.5\%) & 0 (0.0\%) & 2 (0.7\%) & 0 (0.0\%) \\   \cline{2-6}
                                  & Some high school         & 6 (1.6\%) & 3 (3.7\%) & 4 (1.4\%) & 2 (1.9\%) \\ \cline{2-6}
                                  & High school grad         & 39 (10.2\%) & 13 (15.9\%) & 22 (7.9\%) & 17 (16.3\%) \\ \cline{2-6}
                                  & Some college             & 100 (26.2\%) & 28 (34.1\%) & 60 (21.7\%) & 30 (28.8\%) \\ \cline{2-6}
                                  & College grad             & 95 (24.9\%) & 19 (23.2\%) & 75 (27.1\%) & 30 (28.8\%) \\ \cline{2-6}
                                  & Grad/professional school & 126 (33.1\%) & 16 (19.5\%) & 106 (38.3\%) & 23 (22.1\%) \\ \hline\hline
    \end{tabular}
    \caption{\label{tab:demo}Subject demographics of the ECHO RESONANCE cohort.
            "All" refers to the entire cohort.
            "Under 6 months", "Over 18 months", and "Future" refer specifically
            to samples that were used included in the respective subgroup analyses.
    }
\end{table}

@@

\begin{table}[!h]
    \begin{center}
    \begin{tabular}{|r|r|r|r|r|r|r|}
        \hline
        \textbf{Age group}& \multicolumn{2}{c|}{\textbf{0 to 6m}} & \multicolumn{2}{c|}{\textbf{18 to 120m}} \\\hline
        geneset & E.S. & q value & E.S. & q value \\\hline
        Acetate syn. I & -0.165 & 0.284 & 0.192 & \textbf{0.145} \\
        Acetate syn. II & -0.491 & \textbf{0.059} & 0.248 & \textbf{0.15} \\
        Acetate syn. III & -0.409 & \textbf{0.133} & 0.215 & 0.209 \\
        Butyrate syn. II & -0.357 & 0.284 & 0.414 & \textbf{0.178} \\
        Glutamate syn. II & 0.343 & 0.284 & 0.355 & \textbf{0.023} \\
        Propionate deg. I & -0.659 & \textbf{0.029} & 0.154 & \color{red}{\textbf{0.178}} \\\hline
    \end{tabular}
    \caption{\label{tab:fsea} \textbf{Neuroactive microbial gene sets microbial gene sets are associated with cognitive function}.
    Bold text indicates significant (q value \textless 0.2) genesets with a positive or negative
    pseudomedian.}
    \end{center}
\end{table}

@@

\begin{figure}[!h]
    \centering
    \small
    % \includegraphics[width=0.9\textwidth]{assets/Figure1.png}
    \caption{
        \label{fig:1} \textbf{The RESONANCE cohort is a diverse cohort with longitudinal sampling.} (\textbf{A}) Stool samples,
        cognitive assessments and neuroimaging were collected from participants
        at different ages throughout the first years of life.
        (\textbf{B}) Cognitive function scores are assessed by
        different instruments, but may be normalized using full-scale composite scores.
        (\textbf{C-D}) Principal coordinates analysis using Bray-Curtis dissimilarity on
        taxonomic profiles demonstrates high beta diversity, with much of the
        first axis of variation explained by increasing age and alpha diversity.
        Differences in gene function profiles (\textbf{E}) and neuroimaging (PCA based on
        the Euclidean distance of brain region volumes) (\textbf{F}) are likewise
        dominated by changes as children age. (\textbf{G}) Permutation analysis of
        variance (PERMANOVA) of taxonomic profiles, functional profiles
        (annotated as UniRef90s) and neuroimaging, against the metadata of
        interest. Variations in taxonomic and functional profiles explain a
        modest but significant percent of the variation in cognitive development
        in children over 18 months of age. (\textbf{H}) Mantel testing of different
        microbial feature matrices, shows overlapping but distinct patterns of
        variation. Dotted lines in (B) and (C) show 6 and 18 months, which are
        used as cut-offs in some following models.
    }

@@

\begin{figure}[!htb]
    \centering
    \small
    % \includegraphics[width=0.9\textwidth]{assets/Figure4.png}
    \caption{
        \textbf{Microbial features can train RF models to predict MSEL subscores and brain volumes in children over 18 months of age}.
        (\textbf{A}) Test set correlations for RF models predicting MSEL subscale scores,
        with or without demographics included as predictors.
        (\textbf{B}) Important taxa in models predicting MSEL subscale scores.
        (\textbf{C}) Average test-set correlations for prediction of MRI segmentation
        data from microbiome and demographics on select segments after repeated
        cross validation. (\textbf{D}) Heatmap of average individual relative taxa
        importances on each brain segment. Importances are reported as
        proportions relative to the sum of importances for each model - since
        every model is trained with 132 features, and even distribution of
        importance would be 0.75\% for each feature. Segments and taxa ordered
        by HCA on a select list of species with high ranks on average importances,
        and their respective highest-load segments (\textbf{E}).
    }
    \label{fig:4}
\end{figure}

>>>>

\begin{table}[!h]
    \centering
    \begin{tabular}{|r|r|r|r|r|r|}
      \hline\hline
      \textbf{Group}              & \textbf{Subgroup}        & \textbf{All} & \textbf{Under 6 months} & \textbf{Over 18 months} & \textbf{Future} \\\hline
      N subjects                  &                          & 381 & 82 & 277 & 104 \\  \hline
    \multirow{3}{*}{Samples}      & 1                        & 214 (56.2\%) &  &  &  \\ \cline{2-6}
                                  & 2                        & 90 (23.6\%) &  &  &  \\ \cline{2-6}
                                  & \textgreater 2           & 77 (20.2\%) &  &  &  \\ \hline
    \multirow{3}{*}{Age (months)} & min                      & 1.3 & 2.89 & 18.03 & 1.3 \\ \cline{2-6}
                                  & max                      & 119.53 & 5.95 & 119.53 & 11.95 \\ \cline{2-6}
                                  & median                   & 24.62 & 3.64 & 41.59 & 6.08 \\ \hline
    \multirow{2}{*}{Sex}          & F                        & 172 (45.1\%) & 42 (51.2\%) & 122 (44.0\%) & 47 (45.2\%) \\   \cline{2-6}
                                  & M                        & 209 (54.9\%) & 40 (48.8\%) & 155 (56.0\%) & 57 (54.8\%) \\  \hline
    \multirow{6}{*}{Race}         & White                    & 265 (69.6\%) & 48 (58.5\%) & 201 (72.6\%) & 62 (59.6\%) \\   \cline{2-6}
                                  & Black                    & 40 (10.5\%) & 15 (18.3\%) & 21 (7.6\%) & 17 (16.3\%) \\ \cline{2-6}
                                  & Indigenous               & 1 (0.3\%) & 0 (0.0\%) & 1 (0.4\%) & 0 (0.0\%) \\ \cline{2-6}
                                  & Asian                    & 8 (2.1\%) & 2 (2.4\%) & 7 (2.5\%) & 3 (2.9\%) \\ \cline{2-6}
                                  & Mixed                    & 54 (14.2\%) & 14 (17.1\%) & 38 (13.7\%) & 16 (15.4\%) \\ \cline{2-6}
                                  & Other                    & 3 (0.8\%) & 1 (1.2\%) & 3 (1.1\%) & 2 (1.9\%) \\ \hline
    \multirow{6}{*}{Maternal Ed.} & Junior high school       & 2 (0.5\%) & 0 (0.0\%) & 2 (0.7\%) & 0 (0.0\%) \\   \cline{2-6}
                                  & Some high school         & 6 (1.6\%) & 3 (3.7\%) & 4 (1.4\%) & 2 (1.9\%) \\ \cline{2-6}
                                  & High school grad         & 39 (10.2\%) & 13 (15.9\%) & 22 (7.9\%) & 17 (16.3\%) \\ \cline{2-6}
                                  & Some college             & 100 (26.2\%) & 28 (34.1\%) & 60 (21.7\%) & 30 (28.8\%) \\ \cline{2-6}
                                  & College grad             & 95 (24.9\%) & 19 (23.2\%) & 75 (27.1\%) & 30 (28.8\%) \\ \cline{2-6}
                                  & Grad/professional school & 126 (33.1\%) & 16 (19.5\%) & 106 (38.3\%) & 23 (22.1\%) \\ \hline\hline
    \end{tabular}
    \caption{\label{tab:demo}\textbf{Subjects in the ECHO RESONANCE cohort come from a diverse cross-section of Rhode Island families.}
            "All" refers to the entire cohort.
            "Under 6 months", "Over 18 months", and "Future" refer specifically
            to samples that were used included in the respective subgroup analyses.
    }
\end{table}

@@

\begin{table}[!h]
    \begin{center}
    \begin{tabular}{|r|r|r|r|r|r|r|}
        \hline
        \textbf{Age group}& \multicolumn{2}{c|}{\textbf{0 to 6m}} & \multicolumn{2}{c|}{\textbf{18 to 120m}} \\\hline
        geneset & E.S. & q value & E.S. & q value \\\hline
        Acetate syn. I & -0.165 & 0.284 & 0.192 & \textbf{0.145} \\
        Acetate syn. II & -0.491 & \textbf{0.059} & 0.248 & \textbf{0.15} \\
        Acetate syn. III & -0.409 & \textbf{0.133} & 0.215 & 0.209 \\
        Butyrate syn. II & -0.357 & 0.284 & 0.414 & \textbf{0.178} \\
        Glutamate syn. II & 0.343 & 0.284 & 0.355 & \textbf{0.023} \\
        Propionate deg. I & -0.659 & \textbf{0.029} & 0.154 & \textbf{0.178} \\\hline
    \end{tabular}
    \caption{\label{tab:fsea} \textbf{Neuroactive microbial gene sets microbial gene sets are associated with cognitive function}.
    Bold text indicates significant (q value \textless 0.2) genesets with a positive or negative
    pseudomedian.}
    \end{center}
\end{table}

@@

\begin{figure}[!h]
    \centering
    \small
    % \includegraphics[width=0.9\textwidth]{assets/Figure1.png}
    \caption{
        \label{fig:1} \textbf{The RESONANCE cohort is a diverse cohort with accelerated longitudinal sampling.} (\textbf{A}) Stool samples,
        cognitive assessments and neuroimaging were collected from participants
        at different ages throughout the first years of life.
        (\textbf{B}) Cognitive function scores are assessed by
        different instruments, but may be normalized using full-scale composite scores.
        (\textbf{C-D}) Principal coordinates analysis using Bray-Curtis dissimilarity on
        taxonomic profiles demonstrates high beta diversity, with much of the
        first axis of variation explained by increasing age and alpha diversity.
        Differences in gene function profiles (\textbf{E}) and neuroimaging (PCA based on
        the Euclidean distance of brain region volumes) (\textbf{F}) are likewise
        dominated by changes as children age. (\textbf{G}) Permutation analysis of
        variance (PERMANOVA) of taxonomic profiles, functional profiles
        (annotated as UniRef90s) and neuroimaging, against the metadata of
        interest. Variations in taxonomic and functional profiles explain a
        modest but significant percent of the variation in cognitive development
        in children over 18 months of age. (\textbf{H}) Mantel testing of different
        microbial feature matrices, shows overlapping but distinct patterns of
        variation. Dotted lines in (B) and (C) show 6 and 18 months, which are
        used as cut-offs in some following models.
    }

@@

\begin{figure}[!htb]
    \centering
    \small
    % \includegraphics[width=0.9\textwidth]{assets/Figure4.png}
    \caption{
        \textbf{Coupled importance of microbes in neuroanatomy and subscores are consistent with known subcortical associations with cognition.}
        (\textbf{A}) Test set correlations for RF models predicting MSEL subscale scores,
        with or without demographics included as predictors.
        (\textbf{B}) Important taxa in models predicting MSEL subscale scores.
        (\textbf{C}) Average test-set correlations for prediction of MRI segmentation
        data from microbiome and demographics on select segments after repeated
        cross validation. (\textbf{D}) Heatmap of average individual relative taxa
        importances on each brain segment. Importances are reported as
        proportions relative to the sum of importances for each model - since
        every model is trained with 132 features, and even distribution of
        importance would be 0.75\% for each feature. Segments and taxa ordered
        by HCA on a select list of species with high ranks on average importances,
        and their respective highest-load segments (\textbf{E}).
    }
    \label{fig:4}
\end{figure}
